Synthesizing Verified Components for Cyber Assured Systems Engineering \thanks{DISTRIBUTION STATEMENT A. Approved for public release.}

### 2022-03-29 paper/high-assurance-contracts.tex
clarifying exposition of behavior of hardened monitor

\newsavebox{\flt}
\begin{lrbox}{\flt}
  \begin{lstlisting}[style=agree,numbers=left] -- start user
    -- start user definitions
    eq policy : bool =
      WELL_FORMED_AUTOMATION_RESPONSE(Input);
    -- stop user definitions

@@

    eq policy : bool = (rsp => req) -> *\label{line:mon-policy}*
                         (    (isPending => latency < MAX_LATENCY)
                          and (rsp => (req or preIsPending)));
    -- stop user definitions

@@

\subsubsection{Monitor specification}
The monitor policy is given by the \konst{eq}-statements in \linesref{line:mon-assume}{line:mon-policy} of
\figref{fig:monitor}.
\lineref{line:mon-alert}: The \texttt{alert} variable
(distinct from the \texttt{Alert} port) depends not just on the policy
but also on configuration details supplied when the component is
specified.  At configuration time, \texttt{is\_latched} has been set
to \konst{true}, thus making the alert persistent, meaning that once
the alert is raised, it is always raised; otherwise, it is the
complement of the policy value in the current step.

\lineref{line:mon-pre-pending}, \lineref{line:mon-pending}, and \lineref{line:mon-latency}:
The monitor policy is defined by marking when a request is not
satisfied, \texttt{isPending}, and
counting the number of steps between requests, \texttt{latency}.  
The \texttt{isPending} variable is true
whenever there is a request that does not coincide with a
response--to be read as \emph{there is no response since the most recent request without a corresponding response}.  The
\texttt{latency} variable starts at zero, resets on every request, and
otherwise increments by one. 
The \texttt{preIsPending} is needed to make the contract Pascalish since since the current and previous value of \texttt{isPending} are used in the definition of \texttt{policy}.

\lineref{line:mon-policy}: 
The policy makes sure the latency is bounded whenever there is a pending request, and it it makes sure that a response is a result of a request.
The initial step differentiates from later steps in only needing to be sure a response is a result of a request.

\lineref{line:mon-const}: The latency bound for the monitor is defined by the constant
\texttt{MAX\_LATENCY}.  Here the constant
is set to one to be consistent with the system specification from the
previous section.  At time zero the policy is trivial: a response
requires an accompanying request.  After time zero, if a request is
pending in the current time step, then the latency must still be under
the bound, and if there is a response, then it closes a request now or
a pending request from earlier in time.

\lineref{line:mon-assume}: The policy definition only works if there is never more than one
outstanding request at a time; otherwise, the latency counter resets
incorrectly.  That requirement is encoded in the assumption and is the same assumption used for the
system input.  The policy can be written without the assumption,
but the policy is much simpler to write with the assumption.

@@

%%   Precise in this context means that the
%% implementations match the input-to-output relations defined in the
%% specifications.

>>>>

\newsavebox{\flt}
\begin{lrbox}{\flt}
  \begin{lstlisting}[style=agree,numbers=left] -- start user
    -- start user definitions
    eq policy : bool =
      WELL_FORMED_AUTOMATION_RESPONSE(Input);
    -- end user definitions

@@

    eq policy : bool = (rsp => req) -> *\label{line:mon-policy}*
                         (    (isPending => latency < MAX_LATENCY)
                          and (rsp => (req or preIsPending)));
    -- end user definitions

@@

\subsubsection{Monitor specification}

The monitor informally behaves as follows: if the \emph{policy} is
obeyed in the current step, and the alert has not yet been raised,
then the \texttt{Response} input is sent out on \texttt{Output}, and
the \texttt{Alert} output stays
low \lineref{line:mon-alert}. Otherwise \texttt{Alert} goes high and
stays high, and there is nothing sent henceforth on \texttt{Output}.

The \texttt{alert} variable (distinct from the \texttt{Alert} port)
depends not just on the policy but also on the
\texttt{is\_latched} constant, which has been set at configuration time.
Since \texttt{is\_latched} is \konst{true}, the alert is persistent,
meaning that once the alert is raised, it is always raised; otherwise,
it is the complement of the policy value in the current step.

The monitor policy is defined by marking, with variable \texttt{isPending},
when a request is made but not yet satisfied, and by counting the
number of steps (\texttt{latency}) between
requests. See \linesref{line:mon-pre-pending}{line:mon-latency}:
The \texttt{isPending} variable is true whenever there is a request
that does not coincide with a response, \ie, \texttt{isPending} holds
when \emph{there is no response since the most recent request without
a corresponding response}.  The
\texttt{latency} variable starts at zero, resets on every request, and
otherwise increments by one.  Note that the \texttt{preIsPending}
variable is a temporary needed to make the contract \emph{Pascalish};
otherwise, the current and previous value of \texttt{isPending} would
be simulataneously accessed in the definition of \texttt{policy}.

The policy (\lineref{line:mon-policy}) makes sure the latency is
bounded whenever there is a pending request, and it ensures that a
response is a result of a request.  The initial step differentiates
from later steps in only needing to be sure a response is a result of
a request, both happening in the initial step.

The latency bound for the monitor is defined by the constant
\texttt{MAX\_LATENCY} (\lineref{line:mon-const}).  In this example the
constant is set to one to be consistent with the system specification from the
previous section.  At time zero the policy is trivial: a response
requires an accompanying request.  After time zero, if a request is
pending in the current time step, then the latency must still be under
the bound, and if there is a response, then it closes a request now or
a pending request from earlier in time.

The policy definition only works if there is never more than one
outstanding request at a time; otherwise, the latency counter resets
incorrectly.  That requirement is encoded in the assumption
(\lineref{line:mon-assume}) and is the same assumption used for the
system input.  In fact, the policy can be written without the assumption, but
the policy is much simpler to write with the assumption.

@@

### 2022-03-29 paper/synthesis.tex
clarifying exposition of behavior of hardened monitor

\newsavebox{\monFn}
\begin{lrbox}{\monFn}
\begin{lstlisting}[style=myML]
stepFn (Request,Response)
       (req,rsp,current,previous,policy,alert) =
let val stateVars' =
     if !initStep then
        let val req = event(Request)
            val rsp = event(Response)
            val current = (req = rsp)
            val previous = req and not(rsp)
            val policy = current or previous
            val alert = not policy
            val () = (intStep := False)
        in (req,rsp,current,previous,policy,alert)
        end
     else
        let val req = event(Request)
            val rsp = event(Response)
            val current = (req = rsp)
            val previous = pre(req and not rsp) and (not req and rsp)
            val policy = current or previous
            val alert = (is_latched and pre(alert)) or not(policy)
        in (req,rsp,current,previous,policy,alert)
        end
    val (_,rsp',_,_,_,alert') = stateVars'
    val Alert = if alert' then Some () else None
    val Output =
       if alert' then None else
       if rsp'   then Some Response
       else None
in
   (stateVars', (Alert,Output))
end
\end{lstlisting}
\end{lrbox}

>>>>

\newsavebox{\monFn}
\begin{lrbox}{\monFn}
\begin{lstlisting}[style=myML]
fun stepFn (response,request)
    (rsp,req,preIsPending,isPending,latency,policy,alert)
let val stateVars' =
      if !initStep then
      let val rsp = Option.isSome response
          val req = Option.isSome request
          val preIsPending = False
          val isPending = req andalso not rsp
          val latency = 0
          val policy = (if rsp then req else True)
          val alert = not(policy)
          val () = (initStep := False)
      in
        (rsp,req,preIsPending,isPending,latency,policy,alert)
      end
      else
      let ...
          val rsp = Option.isSome response
          val req = Option.isSome request
          val preIsPending = pre isPending
          val isPending = (req andalso not rsp) orelse
                          ((not rsp) andalso pre isPending)
          val latency = (if req then 0 else (pre latency + 1)
          val policy =
            ((if isPending then
                (latency < Defs.max_latency) else True)
              andalso
              (if rsp then
                 (req orelse preIsPending) else True))
          val alert = (Defs.is_latched andalso (pre alert))
                      orelse not(policy)
      in
        (rsp,req,preIsPending,isPending,latency,policy,alert)
      end
   val (rsp',_,_,_,_,_,alert') = stateVars'
   val alertPort = if alert' then Some () else None
   val outputPort =
     if alert' then None else
     if rsp' then Some response
     else None
in
   (stateVars', (alertPort,outputPort))
end
\end{lstlisting}
\end{lrbox}
